PAGAI : a path sensitive static analyser

### 2012-05-18 papers/TAPAS2012/pagai.tex
TAPAS

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
\usepackage[authoryear]{natbib}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
\usepackage{graphicx}
\usepackage{color}

@@

@@

\newcommand{\Nat}{{\mathbb N}}
\newcommand{\Real}{{\mathbb R}}
\def\lastname{Henry}
\begin{document}
\begin{frontmatter}
  \title{PAGAI : a path sensitive static analyser} 
  \author{Julien Henry \thanksref{imag.fr}}
  \address{Universit\'e Joseph Fourier, VERIMAG\\
    Grenoble, France} 
	\author{David Monniaux\thanksref{imag.fr}}
  \address{CNRS, VERIMAG\\
    Grenoble, France} 
	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
  \address{Grenoble-INP, VERIMAG\\
    Grenoble, France}

@@

Exemple du boustrophedon ???
These techniques make a precise analysis of the paths inside the program

@@

	\begin{definition}[Liveness by linearity]
	A variable $v$ is \emph{live by linearity} at a control point $p$ if and
	only if:
	\begin{itemize}
	\item 
	its value is available at $p$, meaning that there is a path from the
	point defining the variable to $p$,
	\item one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live by linearity in $p$.
		\end{itemize}
	\end{itemize}
	\end{definition}

@@

\subsection{Limitations of the tool}

@@

\section{Conclusion and future prospects}
Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain. In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.

A common criticism of analysis techniques based on SMT-solving is that they
do not scale up. Yet, our experiments show that, for numerical properties,
they scale up to the size of typical functions and loops.
It is however
quite certain that, naively applied, they cannot scale to the kind of
programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
of thousands of lines of code in a single loop operating over similar numbers
of remanent variables.
Actually, for such applications, only (quasi-)linear
algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.

We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering. A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.

We focused on numerical properties, because they are supported by easily available abstract libraries. Yet, in most programs, properties of data structures are important for proving interesting properties. Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.

>>>>

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
\usepackage[authoryear]{natbib}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
\usepackage{graphicx}
\usepackage{color}
\usepackage{tikz}

@@

\tikzstyle{arrow}=[->,line width=.05cm,draw=red!90!blue!60!black]

\usetikzlibrary{snakes,arrows,shapes,backgrounds,shadows,automata,patterns}
%\usepgflibrary{snakes}

\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
  			  minimum size=4mm]
\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
  			  minimum size=4mm]
\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
color=blue]
\tikzstyle{line}=[black,thick]

\lstnewenvironment{C}
{\lstset{language=C,
		basicstyle=\ttfamily,
		commentstyle=\textit,
		showstringspaces=false}}
{}

@@

\newcommand{\Nat}{{\mathbb N}}
\newcommand{\Real}{{\mathbb R}}
\def\lastname{Henry, Monniaux \& Moy}
\begin{document}
\begin{frontmatter}
  \title{PAGAI : a path sensitive static analyser} 
  \author{Julien Henry \thanksref{imag.fr}}
  \address{Universit\'e Joseph Fourier, VERIMAG\\
    Grenoble, France} 
	\author{David Monniaux\thanksref{imag.fr}}
  \address{CNRS, VERIMAG\\
    Grenoble, France} 
	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
  \address{Grenoble-INP, VERIMAG\\
    Grenoble, France}

@@

It is well know fact in abstract interpretation that the least upper bound of two or
more abstract values in the abstract domain 
may induce unrecoverable loss of precision. This operation typically occurs with
if-then-else statements and loops.

An example of program where such a loss of precision occurs is depicted in
Figure \ref{example}. In this program, the loop body has two feasible paths that
are executed alternatively, depending on a variable ``phase''. Such kind of
programs often occur in reactive systems. 
Classical abstract interpretation does not prove that $\lstinline|x| < 100$,
because of the abstract union at control point $n_5$. 
To cope with this problem, an idea is compute disjunctive invariants, i.e a
union of values of the abstract domain. However, the number of terms in the
disjunction may grow exponentially with the number of tests in the program.
A second technique is to distinguish all the paths inside the loop, so that we
compute least upper bounds only at the widening points. In this example, if we
distinguish all the 9 paths of the loop, we prove $\lstinline|x| < 100$. Again,
this path enumeration grows exponentially with the number of tests. In order to
avoid this blowup, we use SMT-solving to keep these paths implicit and to
compute them only if needed.

\begin{figure}[!h]
\begin{minipage}[c]{.49\linewidth}
\begin{C}
int x = 0;
int t = 0;
int phase = 0;

while (t < 100) {
   if (phase == 0) 
      x = x+2;
   if (phase == 1) 
      x = x-1;
   phase = 1-phase;
   t++;
}
assert(x <= 100);
\end{C}
\end{minipage}
\begin{minipage}[c]{.49\linewidth}
\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
                    semithick,font=\footnotesize]

	\node[PRstate] (n0) {$n$};
	\node[PRstate] (n00) [below of=n0] {$n_0$};
	\node[PRstate] (nf) [left of=n00] {$n_f$};
	\node[state] (n1) [below of=n00] {$n_1$};
	\node[state] (n3) [below of=n1] {$n_3$};
	\node[state] (n2) [left of=n3] {$n_2$};
	\node[state] (n4) [right of=n3] {$n_4$};
	\node[state] (n5) [below of=n3] {$n_5$};
	\node[state] (n7) [below of=n5] {$n_7$};
	\node[state] (n6) [left of=n7] {$n_6$};
	\node[state] (n8) [right of=n7] {$n_8$};
	\node[state] (n9) [below of=n7] {$n_9$};

  \path [transition] 
		(n0) edge              node  {$\begin{array}{l}
		x \leftarrow 0 \\
		phase \leftarrow 1 
		\end{array}$} (n00);
  \path [transition] 
        (n00)  edge              node {$t>=100$} (nf);
  \path [transition] 
        (n00)  edge              node {$t<100$} (n1);
  \path [transition] 
        (n1) edge			   node [left] {$phase,0?$~ ~ ~ $<$} (n2);
  \path [transition] 
        (n1)  edge              node {$>$} (n3);
  \path [transition] 
        (n1)  edge              node [right] {$=$} (n4);
  \path [transition] 
        (n3) edge              node  {} (n5);
  \path [transition] 
        (n2) edge			   node {} (n5);
  \path [transition] 
        (n4) edge			   node [right] {$x \leftarrow x+2$} (n5);
  \path [transition] 
        (n5) edge			   node [left] {$phase,1?$ ~ ~ ~$<$} (n6);
  \path [transition] 
        (n5) edge			   node {$>$} (n7);
  \path [transition] 
        (n5) edge			   node [right] {$=$} (n8);
  \path [transition] 
        (n6) edge              node {} (n9);
  \path [transition] 
        (n7) edge              node {} (n9);
  \path [transition] 
        (n8) edge              node [right] {$x \leftarrow x-1$} (n9);
  \path [transition] 
        (n9) edge [out=0, in=0, distance=4.3cm] node [right] {
		$\begin{array}{l}
			phase \leftarrow \\
			1-phase \\
			t \leftarrow \\
			t+1
		\end{array}$} (n00);

\end{tikzpicture}
\end{minipage}
\caption{Example of program, where the loop behaviour vary depending on a variable $phase$.}
\label{example}
\end{figure}

@@

	\begin{definition}[Liveness by linearity]
	A variable $v$ is \emph{live by linearity} at a control point $p$ if and
	only if one of these conditions holds:
		\begin{itemize}
		\item $v$ is live in $p$.
		\item There is a variable $v'$, defined as a linear combination of other
		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
		and $v'$ is live by linearity in $p$.
		\end{itemize}
	\end{definition}

@@

\subsection{Use}

PAGAI takes as input an LLVM bitcode file, and outputs an inductive invariant
for each control point in $P_R$ (typically, the widening points).
When a program contains \emph{assert} function call, PAGAI also outputs whether
the statement has been proved.
It is also possible to add some preconditions about the variables, etc, using a
function \emph{assume}.
\subsection{Limitations of the tool}

@@

\section{Conclusion and future work}
